---
title: "Trends in due process hearings over time"
author: "Rebecca Johnson and Rijul Garg"
date: "5/5/2019"
output: html_document
---

```{r}
library(textclean)
library(data.table)
library(janitor)
library(tidyr)
library(ggplot2)
library(rlist)
library(dplyr)
library(devtools)
library("jsonlite")
library(stringr)
library(scales)
library(xtable)
library(stringr)
library(tidyverse)

#devtools::install_github('UrbanInstitute/education-data-package-r')


theme_new <- function(base_size = 24){
  theme_bw(base_size = base_size) %+replace%
    theme(
      panel.grid = element_blank(),   
      panel.border = element_rect(fill = NA, colour = "black", size=1),
      panel.background = element_rect(fill = "white", colour = "black"), 
      strip.background = element_rect(fill = NA),
      axis.text.x = element_text(color = "black"),
      axis.text.y = element_text(color = "black")
      )
}

```


# Step one: load and merge numerator data on complaints by state

General link: https://www2.ed.gov/programs/osepidea/618-data/state-level-data-files/index.html#part-b-menu

## Pull all the years from that site

```{r}
start_fall = 2006
start_spring= 2007
end_fall = 2017
end_spring = 2018

falls = seq(from = start_fall, to = end_fall, by = 1)
springs = seq(from = start_spring, to =  end_spring, by = 1)
spring_suffix = gsub("20", "", springs)
urls_dr = sprintf("https://www2.ed.gov/programs/osepidea/618-data/state-level-data-files/part-b-data/dispute-resolution/bdispres%s-%s.csv",
                  falls,
                  spring_suffix)

### iterate and read in
all_dr_years = list()
for(i in 1:length(urls_dr)){
  one_dr_year = fread(urls_dr[i])
  all_dr_years[[i]] = one_dr_year
}

### remove empty rows
all_dr_years_removeblank = lapply(all_dr_years, function(x){x[rowSums(x == "") == 0]})

### subset to actual data columns
all_dr_years = lapply(all_dr_years_removeblank,
                                function(x){x[2:nrow(x), ]})
all_dr_years_colnames = lapply(all_dr_years_removeblank, 
                               function(x){unlist(x[1, ])})
all_dr_years_colnames_n = lapply(all_dr_years_colnames, 
                                 length)

### clean column names
clean_colnames = lapply(all_dr_years_colnames, function(x) {
  
  x <- gsub("[0-9]+.*$|__",
                            "",
                            unlist(x))
  x <- gsub("[[:punct:]]",
                            "",
                            unlist(x))
  x <- gsub("\\s+",
                            " ",
                            unlist(x))
  x <- gsub("^\\s+|\\s+$",
                            "",
                            unlist(x))
  x <- gsub("\\s+|\\(|\\)", 
                          "_", 
                        unlist(x))
  
  x <- toupper(x)
  return(x)
})

### standardize small differences between 2006/2007 and 2008
clean_colnames[[3]][["V8"]] <- "WSC_Pending_Total"
clean_colnames[[3]][["V9"]] <- "WSC_Pending_a_due_Process_Hearing"
clean_colnames[[3]][["V10"]] <- "WSC_Withdrawn_or_Dismissed_Total"
clean_colnames[[3]][["V14"]] <- "Mediations_Related_to_due_process_Agreements"
clean_colnames[[3]][["V15"]] <- "Mediations_not_Related_to_due_Process"
clean_colnames[[3]][["V16"]] <- "Mediations_not_Related_to_due_process_Agreements"

clean_colnames = lapply(clean_colnames, function(x) { 
  x <- toupper(x)
  return(x)
})

### check remaining differences between data (2006-2008 and 2009-2018)
for(i in 1:length(clean_colnames)){
  print(sprintf("difference between df %s and df %s cols: ",
                i, i+1))
  print(setdiff(unlist(clean_colnames[i]),
                unlist(clean_colnames[i+1])))
  
}

### apply column names and bind standardized data
all_dr_years_firsthree_renamed = lapply(all_dr_years[1:3],
                              setNames,
                            unique(unlist(clean_colnames[1:3])))


all_dr_years_firsthree_bound = do.call(rbind.data.frame,
                                     all_dr_years_firsthree_renamed)

### standardize differences between 2006-2008 and 2009-2018
all_dr_years_firsthree_bound <- all_dr_years_firsthree_bound %>%
  rename(
    WRITTEN_SIGNED_COMPLAINTS_WSC_TOTAL = WRITTEN_SIGNED_COMPLAINTS_TOTAL,
    WSC_REPORTS_WITH_FINDINGS = WSC_REPORTS_WITH_FINDING,
    WSC_REPORTS_WITHIN_TIMELINE = WSC_REPORTS_WITH_TIMELINES,                     
    WSC_REPORTS_WITHIN_EXTENDED_TIMELINES = WSC_REPORTS_WITH_EXTENDED_TIMELINES,  
    WSC_WITHDRAWN_OR_DISMISSED = WSC_WITHDRAWN_OR_DISMISSED_TOTAL,                
    MEDIATIONS_HELD_TOTAL = MEDIATIONS_TOTAL,                                  
    MEDIATIONS_HELD_RELATED_TO_DUE_PROCESS_COMPLAINTS = MEDIATIONS_RELATED_TO_DUE_PROCESS,
    MEDIATION_AGREEMENTS_RELATED_TO_DUE_PROCESS_COMPLAINTS = MEDIATIONS_RELATED_TO_DUE_PROCESS_AGREEMENTS,     
    MEDIATIONS_HELD_NOT_RELATED_TO_DUE_PROCESS_COMPLAINTS = MEDIATIONS_NOT_RELATED_TO_DUE_PROCESS,            
    MEDIATION_AGREEMENTS_NOT_RELATED_TO_DUE_PROCESS_COMPLAINTS = MEDIATIONS_NOT_RELATED_TO_DUE_PROCESS_AGREEMENTS, 
    MEDIATIONS_PENDING = MEDIATIONS_NOT_HELD_INCLUDING_PENDING_TOTAL, # check     
    DUE_PROCESS_COMPLAINTS_DPC_TOTAL = HEARING_REQUESTS_TOTAL,                            
    DPC_RESOLUTION_MEETINGS_TOTAL = HR_RESOLUTION_SESSIONS_TOTAL,                    
    DPC_RESOLUTION_MEETINGS_WRITTEN_SETTLEMENT_AGREEMENTS = HR_RESOLUTIONS_SETTLEMENT_AGREEMENTS,             
    DPC_HEARINGS_FULLY_ADJUDICATED_TOTAL = HEARINGS_FULLY_ADJUDICATED_TOTAL,                  
    DPC_WRITTEN_DECISIONS_WITHIN_TIMELINE = HEARINGS_DECISIONS_WI_TIMELINE,             
    DPC_WRITTEN_DECISIONS_WITHIN_EXTENDED_TIMELINES = HEARINGS_DECISIONS_WI_EXTENDED_TIMELINE,          
    DPC_WITHDRAWN_OR_DISMISSED = HR_RESOLVED_WITHOUT_A_HEARING_TOTAL, # check              
    EXPEDITED_DUE_PROCESS_COMPLAINTS_EDPC_TOTAL = EXPEDITED_HEARING_REQUESTS_TOTAL,                  
    EDPC_RESULTED_IN_A_RESOLUTION_MEETING_TOTAL =  EHR_RESOLUTION_SESSIONS_TOTAL,                     
    EDPC_RESOLUTION_MEETINGS_WRITTEN_SETTLEMENT_AGREEMENTS = RESOLUTION_SESSIONS_SETTLEMENT_AGREEMENTS, 
    EDPC_EXPEDITED_HEARINGS_FULLY_ADJUDICATED_TOTAL = EHR_EXPEDITED_HEARINGS_FULLY_ADJUDICATED_TOTAL, 
    EDPC_EXPEDITED_HEARINGS_CHANGE_OF_PLACEMENT_ORDERED = EHR_EXPEDITED_HEARINGS_CHANGE_OF_PLACEMENT_ORDERED
  )

### bind all data
all_dr_years_4through12_renamed = lapply(all_dr_years[4:12],
                              setNames,
                          unique(unlist(clean_colnames[4:12])))
all_dr_years_4through12_bound = do.call(rbind.data.frame,
                                all_dr_years_4through12_renamed)
table(all_dr_years_4through12_bound$Year)
table(all_dr_years_firsthree_bound$Year)

df_analytic = rbind(all_dr_years_firsthree_bound, all_dr_years_4through12_bound, fill=TRUE) 

### clean bound data
df_analytic$STATE <-gsub("[[:punct:]]|\\\xa0",
                            "",
                            df_analytic$STATE)
df_analytic$YEAR <-gsub("-.*$",
                            "",
                            df_analytic$YEAR)

df_analytic$DUE_PROCESS_COMPLAINTS_DPC_TOTAL <- as.numeric(gsub(",","",df_analytic$DUE_PROCESS_COMPLAINTS_DPC_TOTAL))

## check number of states
nonstates_dispute = paste(gsub("^\\s+|\\s+$", "", 
              toupper(unique(c(grep("Republic", unique(df_analytic$STATE), value = TRUE, ignore.case = TRUE), 
              grep("Outlying", unique(df_analytic$STATE), value = TRUE, ignore.case = TRUE),
              grep("Puerto|Federated|Marianas|BIE|Samoa|50 states|Bureau|Guam|Virgin Island|Indian Education|Marshall", unique(df_analytic$STATE), 
                   value = TRUE, ignore.case = TRUE))))), collapse = "|") 

df_analyt_cleanstates = df_analytic %>%
              mutate(STATE = toupper(STATE)) %>%
              filter(!grepl(nonstates_dispute, STATE))

stopifnot(length(unique(df_analyt_cleanstates$STATE)) == 51)
```

## Write complaints data to file

```{r}

### set working directory so that it's where
### current file is stored
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))


dispute_analytic <- df_analytic
fwrite(dispute_analytic,
"../intermediate_objects/dph_hearings_20062018.csv")



```

# Step two: load and merge data on denominator (total count of students and count of students with ieps)

## Denom 1: total count of students

Original data source: NCES, Common core data, elsi table generator: https://nces.ed.gov/ccd/elsi/tableGenerator.aspx


```{r}
totalenroll_raw = read.csv("../raw_input/allstates/elsi_enrollment_removeheaderfooter.csv")

### reorganize data to long format and select total students all grades excludes AE as the correct count
totalenroll_long <- gather(totalenroll_raw, Year, "Student_Count", Total.Students..State..2017.18:Total.Students.All.Grades..Excludes.AE...Public.School..2006.07, factor_key=TRUE)

### clean data values and column names
totalenroll_analytic <- totalenroll_long %>%
  dplyr::select(State = State.Name, Year, "Student_Count") %>%
  filter(grepl('All', Year)) %>%
  mutate(Year = str_extract(Year, "(\\d+)+")) %>%
  rename_all(toupper) %>%
  mutate_if(is.factor, as.character) %>%
  mutate(STATE = tools::toTitleCase(tolower(STATE)))
```

## Write student enrollment data to file

```{r}

### set working directory so that it's where current file is stored
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

fwrite(totalenroll_analytic,
"../intermediate_objects/studentenrollment_allstates_20062018.csv")


```

## Denom 2: count of students with IEP (IDEA part B)

Get from same site as complaints data since not in NCES data

```{r}
### link for sy 2005-2006 to 2011-2012
baselink_earlier_root = "https://www2.ed.gov/programs/osepidea/618-data/state-level-data-files/part-b-data/child-count/bchildcount"
years_earlier = seq(from = 2005, to = 2006)
baselink_earlier_full = sprintf("%s%s.csv",
                                baselink_earlier_root,
                                years_earlier)
all_earlier_iep = lapply(baselink_earlier_full,
                         function(x) fread(x))

years_middle = seq(from = 2007, to = 2011)
baselink_middle_full = sprintf("%s%s.csv",
                                baselink_earlier_root,
                                years_middle)
all_middle_iep = lapply(baselink_middle_full,
                         function(x) fread(x))

### clean headers of data
head(all_earlier_iep[[1]])

clean_earlier_iep = lapply(all_earlier_iep,
                           function(x) { 
  x <- x %>%
    row_to_names(row_number = 5)
  
  return(x)
})


clean_middle_iep = lapply(all_middle_iep,
                           function(x) { 
  x <- x %>%
    row_to_names(row_number = 4)
  
  return(x)
})

### clean data and select pertinent columns of 2005-2011
filtered_earlier_iep = lapply(clean_earlier_iep, 
                               function(x) {
                                 x <- x %>%
                                   filter(Disability == "All disabilities") %>%
                                   dplyr::select(Year, State, "Age 3 to 21" = "Age 3-21 Totals")
                                 return(x)
                               })

filtered_middle_iep = lapply(clean_middle_iep, 
                               function(x) {
                                 x <- x %>%
                                   filter(Disability == "All disabilities")
                                 return(x)
                               })

names(filtered_middle_iep[[1]]) <- gsub("[\r\n]",
                            "", names(filtered_middle_iep[[1]]))
filtered_first_middle_iep <- filtered_middle_iep[[1]] %>%
  rename("Age 3 to 21" = "Ages3_21Total ") %>%
  dplyr::select(Year, State, "Age 3 to 21")

filtered_rest_middle_iep <- lapply(filtered_middle_iep[2:5], 
                               function(x) {
                                 x <- x %>%
                                   dplyr::select(Year, State, "Age 3 to 21")
                                 return(x)
                               })

filtered_earlier_iep_bound <- do.call(rbind.data.frame,
                                filtered_earlier_iep)

filtered_rest_middle_iep_bound <- do.call(rbind.data.frame,
                                filtered_rest_middle_iep)

filtered_earlier_middle_iep_bound = rbind(filtered_earlier_iep_bound, filtered_first_middle_iep, filtered_rest_middle_iep_bound)

## repeat for later links
baselink_later_root = "https://www2.ed.gov/programs/osepidea/618-data/state-level-data-files/part-b-data/child-count-and-educational-environments/bchildcountandedenvironments"
years_later = seq(from = 2012, to = 2017)

baselink_later_full_1 = sprintf("%s%s.csv",
                                baselink_later_root,
                                years_later[years_later < 2017])
baselink_later_full_2 = sprintf("%s%s-%s.csv",
                                baselink_later_root,
                                years_later[years_later >= 2017],
                               gsub("20", 
                                    "", 
                                    years_later[years_later >= 2017] +1))

all_later_iep = lapply(c(baselink_later_full_1,
                         baselink_later_full_2),
                         function(x) fread(x))

### clean headers of data
clean_later_iep = lapply(all_later_iep,
                           function(x) { 
  x <- x %>%
    row_to_names(row_number = 5) %>%
    clean_names()
  return(x)
})

clean_2016_iep <- all_later_iep[[5]] %>%
    row_to_names(row_number = 4) %>%
    clean_names() %>%
    dplyr::select(year, state_name, sea_education_environment, sea_disability_category, age_3_to_5, ages_6_21)

clean_later_iep[[5]] <- clean_2016_iep
filtered_later_iep = lapply(clean_later_iep,
                           function(x) { 
  x <- x %>%
    dplyr::select(year, state_name, sea_education_environment, sea_disability_category, age_3_to_5, ages_6_21) %>%
    filter(sea_disability_category == "All Disabilities") %>%
    filter(grepl("Total",sea_education_environment))%>%
    mutate(age_3_to_5=replace(age_3_to_5, age_3_to_5 == "-", 0)) %>%
    mutate(ages_6_21=replace(ages_6_21, ages_6_21 == "-", 0)) %>%
    group_by(year, state_name) %>%
    summarise("Age 3 to 21" = sum(as.numeric(age_3_to_5), as.numeric(ages_6_21))) %>%
    rename(State = state_name, Year = year)
  return(x)
})

filtered_later_iep_bound <- do.call(rbind.data.frame,
                                filtered_later_iep)

iep_init = rbind(data.frame(filtered_earlier_middle_iep_bound), data.frame(filtered_later_iep_bound))

nonstates = paste(toupper(unique(c(grep("Republic", unique(iep_init$State), value = TRUE, ignore.case = TRUE), 
              grep("Outlying", unique(iep_init$State), value = TRUE, ignore.case = TRUE),
              grep("Puerto|Federated|Marianas|BIE|Samoa|50 states|Bureau|Guam|Virgin Island", unique(iep_init$State), 
                   value = TRUE, ignore.case = TRUE)))), collapse = "|") 

iep_analytic <- iep_init %>%
  rename_all(toupper) %>%
  mutate(STATE = toupper(STATE), # since half states are lowercase; others are uppercase
    AGE_3_21_TOTAL = as.numeric(gsub(",", "", AGE.3.TO.21))) %>%
  dplyr::select(YEAR, STATE, AGE_3_21_TOTAL) %>%
  filter(!grepl(nonstates, STATE)) #filtering out those outside 50 states = DC


## check to see if # of unique states equal to 51 
## 50 states + DC
stopifnot(length(unique(iep_analytic$STATE)) == 51)
```


## Write to intermed analytic to file

```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

fwrite(iep_analytic,
"../intermediate_objects/iepcounts_allstates_20052018.csv")

```

# Step three: construct two rates and summarize 

- Count of due process hearing *requests* by state divided by total student enrollment. You'll want to look at the documentation for the complaints data to get the right var/you can ping me on slack when the data is clean, but I think the varname is something like DPCTotal (which should be a higher count than the variable DPC adjudicated which references filings that go to a hearing)
- Count of due process hearing *requests* by state divided by total number of students aged 3-21 with IEP

- Show DC first compared to median of other states
- Then compare to a few other high complaint east coast states (e.g., NY)

```{r}
## before merges, check overlap



dph_tse <- merge(dispute_analytic, totalenroll_analytic, by=c('YEAR','STATE'),
                 all.x = TRUE)

dph_tse$STUDENT_COUNT <- dph_tse$STUDENT_COUNT %>% as.numeric
dph_tse$dph_tse_rate <- dph_tse$DUE_PROCESS_COMPLAINTS_DPC_TOTAL / dph_tse$STUDENT_COUNT * 100


dph_iep <- merge(dph_tse, iep_analytic, by=c('YEAR','STATE'))

dph_iep$AGE_3_21_TOTAL <- as.numeric(gsub(",","",dph_iep$AGE_3_21_TOTAL))
dph_iep$dph_iep_rate <- dph_iep$DUE_PROCESS_COMPLAINTS_DPC_TOTAL / dph_iep$AGE_3_21_TOTAL * 100

median_dph <- dph_iep %>%
  filter(STATE != "District of Columbia") %>%
  group_by(YEAR) %>%
  summarize(median_tse_rate = median(dph_tse_rate, na.rm = TRUE), median_iep_rate = median(dph_iep_rate, na.rm = TRUE))

dph_states <- dph_iep %>%
  filter(STATE == "District of Columbia" | STATE == "New York" | STATE == "New Jersey" | STATE == "Massachusetts")
  
dph_dc <- dph_iep %>%
  filter(STATE == "District of Columbia")
dph_analytic <- merge(dph_dc,median_dph,by='YEAR')

ggplot(data=dph_analytic, aes(x=YEAR,  group=1)) +
  geom_line(aes(y=dph_tse_rate), color = 'blue') +
  geom_line(aes(y=dph_iep_rate), color = 'blue', linetype = "dashed") +
  geom_line(aes(y=median_tse_rate)) +
  geom_line(aes(y=median_iep_rate), linetype = "dashed") +
  scale_x_discrete(name = "Year") +
  scale_y_continuous(name = "Filings per Student (%)", limits = c(0, 31),breaks = seq(0, 31, 5)) +
  theme_bw() +
  ggtitle("District of Columbia Rates vs. Other States Median Rates")

ggplot(data=dph_states, aes(x=YEAR,  group=STATE)) +
  geom_line(aes(y=dph_tse_rate, color=STATE)) +
  geom_line(aes(y=dph_iep_rate, color=STATE)) +
  scale_x_discrete(name = "Year") +
  scale_y_continuous(name = "Filings per Student (%)", limits = c(0, 31),breaks = seq(0, 31, 5)) +
  theme_bw() +
  ggtitle("District of Columbia Rates vs. High Frequency East Coast State Rates")


```
## Older code that might be relevant but refers to older data / varnames

Might be useful but doesn't run as is due to changes in the input data!


```{r}


## merge using state name
colnames(compare_states) = gsub("\\s+", "_", colnames(compare_states))
compare_states = compare_states %>%
            rename(total_students = `Total_Students_[State]_2013-14`,
                   total_districts = `Total_Number_Operational_School_Districts_[District]_2013-14`)
unique(compare_states$State_Name)
unique(dph$State)


dph_mainyear = dph %>%
      filter(Year == "2016-2017") %>%
      mutate(dph_filed = as.numeric(gsub(",", "", as.character(`Due_Process_ComplaintsDPCTotal`))),
             dph_hearings = as.numeric(gsub(",", "", 
                              as.character(`DPC_Hearingsfully_adjudicatedTotal`)))) %>%
      dplyr::select(State, dph_filed, dph_hearings) %>%
      mutate(State_Name = toupper(State)) %>%
      dplyr::select(-State) 
      

## merge

compare_w_dph = merge(compare_states,
                      dph_mainyear,
                      by = "State_Name",
                      all.x = TRUE) %>%
            left_join(ieps %>% dplyr::select(State_Name, ncesiepcount),
                      on = "State_Name") %>%
            mutate(ncesiepcount = ifelse(State_Name == "DISTRICT OF COLUMBIA",
                                         10644, ncesiepcount)) %>%
            mutate(filings_perstudent = (dph_filed/total_students)*10000,
                   hearings_perstudent = (dph_hearings/total_students)*10000,
                   ieps_perstudent= (ncesiepcount/total_students)*100,
                   students_per_district = total_students/total_districts,
                   state_category = case_when(State_Name %in% c("DISTRICT OF COLUMBIA",
                                                                "TEXAS",
                                                                "OHIO") ~ "Case study",
                                    State_Name %in% states_obtained ~ "Filings obtained",
                                    TRUE ~ "Filings not obtained"),
                   state_title = str_to_title(State_Name),
                   hearing_per_filing = ifelse(dph_filed == 0,
                                          0, dph_hearings/dph_filed))

ggplot(compare_w_dph, aes(x = reorder(state_title, filings_perstudent), 
                          y = filings_perstudent,
                          color = state_category)) +
  geom_point(size = 4) +
  ylab("Due process filings per\n10,000 students") +
  theme_new(base_size = 24) +
  theme(axis.text.y = element_text(size = 10),
        legend.position = c(0.7, 0.2),
        legend.background = element_blank()) +
  scale_color_manual(values = c("#FDE725FF", 
                                "wheat4",
                              "#238A8DFF")) +
  coord_flip() +
  labs(color = "") +
  xlab("")

ggsave("../output/dph_rate.pdf",
       plot = last_plot(),
       width = 9,
       height = 12,
       device = "pdf")


## create dc specific plot
focus_states = compare_w_dph %>% filter(State_Name %in% c("DISTRICT OF COLUMBIA", "VIRGINIA", "MARYLAND", "NEW YORK"))
national_median = quantile(compare_w_dph$filings_perstudent,
                           probs = 0.5)
focus_states_forplot = rbind.data.frame(focus_states %>%
                                      dplyr::select(state_title,
                                                    filings_perstudent,
                                                    state_category),
                          data.frame(state_title = "National\nmedian",
                                     filings_perstudent = national_median,
                                     state_category = "Filings obtained")) %>%
                          mutate(clean_state = 
                                      case_when(state_title == "District Of Columbia" ~
                                                "DC",
                                                TRUE ~ state_title),
                                 clean_category = 
                                  case_when(clean_state == "DC" ~ "Case study",
                                            TRUE ~ "Other"))
ggplot(focus_states_forplot, aes(x = reorder(clean_state, filings_perstudent), 
                          y = filings_perstudent,
                          fill = clean_category)) +
  geom_bar(stat = "identity") +
  ylab("Due process filings per\n10,000 students") +
  theme_new(base_size = 24) +
  theme(legend.position = c(0.7, 0.2),
        legend.background = element_blank()) +
  scale_fill_manual(values = c("Case study" = "#132460",
                                "Other" = "#999999")) +
  coord_flip() +
  labs(color = "") +
  xlab("") +
  geom_label(aes(x = reorder(clean_state, filings_perstudent),
                 y = filings_perstudent,
                  label = round(filings_perstudent, 2)),
             fill = "white", size = 14) +
  guides(fill = FALSE)  +
  ylim(-5, 50)

ggsave("../output/dph_rate_dcfocus.jpg",
       plot = last_plot(),
       width = 9,
       height = 12,
       device = "jpg")
  

ggplot(compare_w_dph %>%
      filter(state_title != "Hawaii"),
      aes(x = reorder(state_title, students_per_district), 
                          y = students_per_district,
                          color = state_category)) +
  geom_point(size = 4) +
  ylab("Total students divided by total\nnumber of districts") +
  theme_new(base_size = 24) +
  scale_y_continuous(breaks = pretty_breaks(n = 10)) +
  theme(axis.text.y = element_text(size = 10),
        legend.position = c(0.7, 0.2),
        axis.text.x = element_text(angle = 90),
        legend.background = element_blank()) +
  scale_color_manual(values = c("#FDE725FF", 
                                "wheat4",
                              "#238A8DFF")) +
  coord_flip() +
  labs(color = "") +
  xlab("")

ggsave("../output/dist_consolidation.pdf",
       plot = last_plot(),
       width = 9,
       height = 12,
       device = "pdf")

ggplot(compare_w_dph, aes(x = reorder(state_title, ieps_perstudent), 
                          y = ieps_perstudent,
                          color = state_category)) +
  geom_point(size = 4) +
  ylab("Percent of students who have\nIEPs") +
  theme_new(base_size = 24) +
  scale_y_continuous(breaks = pretty_breaks(n = 10)) +
  theme(axis.text.y = element_text(size = 10),
        legend.position = c(0.3, 0.8),
        legend.background = element_blank()) +
  scale_color_manual(values = c("#FDE725FF", 
                                "wheat4",
                              "#238A8DFF")) +
  coord_flip() +
  labs(color = "") +
  xlab("")

ggsave("../output/iep_rate.pdf",
       plot = last_plot(),
       width = 9,
       height = 12,
       device = "pdf")







## 
dph_states_wdata = dph %>%
          filter(State %in% filings_progress$state)



dist = read.csv("../data/allstates/district_consolidation.csv") %>%
      mutate(n_students = )

dph_1314 = dph %>%
  filter(Year == "2013-2014") %>%
  left_join(statedisp %>% filter(springschoolyear == 2014) %>% %>% mutate(state = as.character(state)) %>%
                    dplyr::select(state, ncesiepcount),
                    by = "state")




filings_progress = filings_progress %>%
          left_join(dph %>% 
                      filter(Year == "2012-2013") %>%
                      dplyr::select(State, Due_Process_ComplaintsDPCTotal),
                    by = c("state" = "State")) %>%
          left_join(statedisp %>% filter(springschoolyear == 2013) %>% mutate(state = as.character(state)) %>%
                    dplyr::select(state, ncesiepcount),
                    by = "state") %>%
        left_join(dist, by = "state") %>%
          mutate(n_districts = as.numeric(gsub(",", "", n_districts)), 
                 students_per_district = as.numeric(gsub(",", "", avg_per_district)),
                 complaints_total = as.numeric(Due_Process_ComplaintsDPCTotal),
                 ieps_total = ifelse(state != "District of Columbia",
                                  as.numeric(ncesiepcount), 
                                  11000),
              complaints_per_100iep = (complaints_total/ieps_total)*100) %>%
            dplyr::select(-avg_per_district, -excluded_districts, -n_schools,
                          -Due_Process_ComplaintsDPCTotal, -ncesiepcount, -n_students)
cols_select = c("state", "years", "format", "n_districts",
                "students_per_district",
                "complaints_total", "ieps_total", "complaints_per_100iep")
print(xtable(filings_progress[, cols_select]), include.rownames = FALSE)

```

