---
title: "Trends in due process hearings over time"
author: "Rebecca Johnson and Rijul Garg"
date: "5/5/2019"
output: html_document
---

```{r}
library(textclean)
library(data.table)
library(janitor)
library(tidyr)
library(ggplot2)
library(rlist)
library(dplyr)
library(devtools)
library("jsonlite")
library(stringr)
library(scales)
library(xtable)
library(stringr)
library(tidyverse)
library(stringr)



theme_new <- function(base_size = 24){
  theme_bw(base_size = base_size) %+replace%
    theme(
      panel.grid = element_blank(),   
      panel.border = element_rect(fill = NA, colour = "black", size=1),
      panel.background = element_rect(fill = "white", colour = "black"), 
      strip.background = element_rect(fill = NA),
      axis.text.x = element_text(color = "black"),
      axis.text.y = element_text(color = "black")
      )
}

```


# Step one: load and merge numerator data on complaints by state

General link: https://www2.ed.gov/programs/osepidea/618-data/state-level-data-files/index.html#part-b-menu

## Pull all the years from that site

```{r}
start_fall = 2006
start_spring= 2007
end_fall = 2017
end_spring = 2018

falls = seq(from = start_fall, to = end_fall, by = 1)
springs = seq(from = start_spring, to =  end_spring, by = 1)
spring_suffix = gsub("20", "", springs)
urls_dr = sprintf("https://www2.ed.gov/programs/osepidea/618-data/state-level-data-files/part-b-data/dispute-resolution/bdispres%s-%s.csv",
                  falls,
                  spring_suffix)

### iterate and read in
all_dr_years = list()
for(i in 1:length(urls_dr)){
  one_dr_year = fread(urls_dr[i])
  all_dr_years[[i]] = one_dr_year
}

### remove empty rows
all_dr_years_removeblank = lapply(all_dr_years, function(x){x[rowSums(x == "") == 0]})

### subset to actual data columns
all_dr_years = lapply(all_dr_years_removeblank,
                                function(x){x[2:nrow(x), ]})
all_dr_years_colnames = lapply(all_dr_years_removeblank, 
                               function(x){unlist(x[1, ])})
all_dr_years_colnames_n = lapply(all_dr_years_colnames, 
                                 length)

### clean column names
clean_colnames = lapply(all_dr_years_colnames, function(x) {
  
  x <- gsub("[0-9]+.*$|__",
                            "",
                            unlist(x))
  x <- gsub("[[:punct:]]",
                            "",
                            unlist(x))
  x <- gsub("\\s+",
                            " ",
                            unlist(x))
  x <- gsub("^\\s+|\\s+$",
                            "",
                            unlist(x))
  x <- gsub("\\s+|\\(|\\)", 
                          "_", 
                        unlist(x))
  
  x <- toupper(x)
  return(x)
})

### standardize small differences between 2006/2007 and 2008
clean_colnames[[3]][["V8"]] <- "WSC_Pending_Total"
clean_colnames[[3]][["V9"]] <- "WSC_Pending_a_due_Process_Hearing"
clean_colnames[[3]][["V10"]] <- "WSC_Withdrawn_or_Dismissed_Total"
clean_colnames[[3]][["V14"]] <- "Mediations_Related_to_due_process_Agreements"
clean_colnames[[3]][["V15"]] <- "Mediations_not_Related_to_due_Process"
clean_colnames[[3]][["V16"]] <- "Mediations_not_Related_to_due_process_Agreements"

clean_colnames = lapply(clean_colnames, function(x) { 
  x <- toupper(x)
  return(x)
})

### check remaining differences between data (2006-2008 and 2009-2018)
for(i in 1:length(clean_colnames)){
  print(sprintf("difference between df %s and df %s cols: ",
                i, i+1))
  print(setdiff(unlist(clean_colnames[i]),
                unlist(clean_colnames[i+1])))
  
}

### apply column names and bind standardized data
all_dr_years_firsthree_renamed = lapply(all_dr_years[1:3],
                              setNames,
                            unique(unlist(clean_colnames[1:3])))


all_dr_years_firsthree_bound = do.call(rbind.data.frame,
                                     all_dr_years_firsthree_renamed)

### standardize differences between 2006-2008 and 2009-2018
all_dr_years_firsthree_bound <- all_dr_years_firsthree_bound %>%
  rename(
    WRITTEN_SIGNED_COMPLAINTS_WSC_TOTAL = WRITTEN_SIGNED_COMPLAINTS_TOTAL,
    WSC_REPORTS_WITH_FINDINGS = WSC_REPORTS_WITH_FINDING,
    WSC_REPORTS_WITHIN_TIMELINE = WSC_REPORTS_WITH_TIMELINES,                     
    WSC_REPORTS_WITHIN_EXTENDED_TIMELINES = WSC_REPORTS_WITH_EXTENDED_TIMELINES,  
    WSC_WITHDRAWN_OR_DISMISSED = WSC_WITHDRAWN_OR_DISMISSED_TOTAL,                
    MEDIATIONS_HELD_TOTAL = MEDIATIONS_TOTAL,                                  
    MEDIATIONS_HELD_RELATED_TO_DUE_PROCESS_COMPLAINTS = MEDIATIONS_RELATED_TO_DUE_PROCESS,
    MEDIATION_AGREEMENTS_RELATED_TO_DUE_PROCESS_COMPLAINTS = MEDIATIONS_RELATED_TO_DUE_PROCESS_AGREEMENTS,     
    MEDIATIONS_HELD_NOT_RELATED_TO_DUE_PROCESS_COMPLAINTS = MEDIATIONS_NOT_RELATED_TO_DUE_PROCESS,            
    MEDIATION_AGREEMENTS_NOT_RELATED_TO_DUE_PROCESS_COMPLAINTS = MEDIATIONS_NOT_RELATED_TO_DUE_PROCESS_AGREEMENTS, 
    MEDIATIONS_PENDING = MEDIATIONS_NOT_HELD_INCLUDING_PENDING_TOTAL, # check     
    DUE_PROCESS_COMPLAINTS_DPC_TOTAL = HEARING_REQUESTS_TOTAL,                            
    DPC_RESOLUTION_MEETINGS_TOTAL = HR_RESOLUTION_SESSIONS_TOTAL,                    
    DPC_RESOLUTION_MEETINGS_WRITTEN_SETTLEMENT_AGREEMENTS = HR_RESOLUTIONS_SETTLEMENT_AGREEMENTS,             
    DPC_HEARINGS_FULLY_ADJUDICATED_TOTAL = HEARINGS_FULLY_ADJUDICATED_TOTAL,                  
    DPC_WRITTEN_DECISIONS_WITHIN_TIMELINE = HEARINGS_DECISIONS_WI_TIMELINE,             
    DPC_WRITTEN_DECISIONS_WITHIN_EXTENDED_TIMELINES = HEARINGS_DECISIONS_WI_EXTENDED_TIMELINE,          
    DPC_WITHDRAWN_OR_DISMISSED = HR_RESOLVED_WITHOUT_A_HEARING_TOTAL, # check              
    EXPEDITED_DUE_PROCESS_COMPLAINTS_EDPC_TOTAL = EXPEDITED_HEARING_REQUESTS_TOTAL,                  
    EDPC_RESULTED_IN_A_RESOLUTION_MEETING_TOTAL =  EHR_RESOLUTION_SESSIONS_TOTAL,                     
    EDPC_RESOLUTION_MEETINGS_WRITTEN_SETTLEMENT_AGREEMENTS = RESOLUTION_SESSIONS_SETTLEMENT_AGREEMENTS, 
    EDPC_EXPEDITED_HEARINGS_FULLY_ADJUDICATED_TOTAL = EHR_EXPEDITED_HEARINGS_FULLY_ADJUDICATED_TOTAL, 
    EDPC_EXPEDITED_HEARINGS_CHANGE_OF_PLACEMENT_ORDERED = EHR_EXPEDITED_HEARINGS_CHANGE_OF_PLACEMENT_ORDERED
  )

### bind all data
all_dr_years_4through12_renamed = lapply(all_dr_years[4:12],
                              setNames,
                          unique(unlist(clean_colnames[4:12])))
all_dr_years_4through12_bound = do.call(rbind.data.frame,
                                all_dr_years_4through12_renamed)
table(all_dr_years_4through12_bound$Year)
table(all_dr_years_firsthree_bound$Year)

df_analytic = rbind(all_dr_years_firsthree_bound, all_dr_years_4through12_bound, fill=TRUE) 

### clean bound data
df_analytic$STATE <-gsub("[[:punct:]]|\\\xa0",
                            "",
                            df_analytic$STATE)
df_analytic$YEAR <-gsub("-.*$",
                            "",
                            df_analytic$YEAR)


## convert char with "," to numeric
make_num <- function(one_var){
  
  return(as.numeric(gsub(",", "", one_var)))
}

df_analytic <- df_analytic %>%
        mutate(DUE_PROCESS_COMPLAINTS_DPC_TOTAL = make_num(DUE_PROCESS_COMPLAINTS_DPC_TOTAL),
               EXPEDITED_DUE_PROCESS_COMPLAINTS_EDPC_TOTAL = make_num(EXPEDITED_DUE_PROCESS_COMPLAINTS_EDPC_TOTAL),
               MEDIATION_REQUESTS_TOTAL = make_num(MEDIATION_REQUESTS_TOTAL),
               WRITTEN_SIGNED_COMPLAINTS_WSC_TOTAL = make_num(WRITTEN_SIGNED_COMPLAINTS_WSC_TOTAL))

## check number of states
nonstates_dispute = paste(gsub("^\\s+|\\s+$", "", 
              toupper(unique(c(grep("Republic", unique(df_analytic$STATE), value = TRUE, ignore.case = TRUE), 
              grep("Outlying", unique(df_analytic$STATE), value = TRUE, ignore.case = TRUE),
              grep("Puerto|Federated|Marianas|BIE|Samoa|50 states|Bureau|Guam|Virgin Island|Indian Education|Marshall", unique(df_analytic$STATE), 
                   value = TRUE, ignore.case = TRUE))))), collapse = "|") 

df_analyt_cleanstates = df_analytic %>%
              mutate(STATE = toupper(STATE),
                     FALL_SY = gsub("\\-.*", "", YEAR)) %>%
              filter(!grepl(nonstates_dispute, STATE))

stopifnot(length(unique(df_analyt_cleanstates$STATE)) == 51)
```

## Write complaints data to file

```{r}

### set working directory so that it's where
### current file is stored
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))


dispute_analytic <- df_analyt_cleanstates
fwrite(dispute_analytic,
"../intermediate_objects/dph_hearings_20062018.csv")



```

# Step two: load and merge data on denominator (total count of students and count of students with ieps)

## Denom 1: total count of students

Original data source: NCES, Common core data, elsi table generator: https://nces.ed.gov/ccd/elsi/tableGenerator.aspx


```{r}
totalenroll_raw = read.csv("../raw_input/allstates/elsi_enrollment_removeheaderfooter.csv")

### reorganize data to long format and select total students all grades excludes AE as the correct count
totalenroll_long <- gather(totalenroll_raw, Year, "Student_Count", Total.Students..State..2017.18:Total.Students.All.Grades..Excludes.AE...Public.School..2006.07, factor_key=TRUE)

### clean data values and column names
totalenroll_analytic <- totalenroll_long %>%
  dplyr::select(State = State.Name, Year, "Student_Count") %>%
  filter(grepl('All', Year)) %>%
  mutate(Year = str_extract(Year, "(\\d+)+")) %>%
  rename_all(toupper) %>%
  mutate_if(is.factor, as.character) %>%
  mutate(STATE = toupper(STATE))
```

## Write student enrollment data to file

```{r}

### set working directory so that it's where current file is stored
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

fwrite(totalenroll_analytic,
"../intermediate_objects/studentenrollment_allstates_20062018.csv")


```

## Denom 2: count of students with IEP (IDEA part B)

Get from same site as complaints data since not in NCES data

```{r}
### link for sy 2005-2006 to 2011-2012
baselink_earlier_root = "https://www2.ed.gov/programs/osepidea/618-data/state-level-data-files/part-b-data/child-count/bchildcount"
years_earlier = seq(from = 2005, to = 2006)
baselink_earlier_full = sprintf("%s%s.csv",
                                baselink_earlier_root,
                                years_earlier)
all_earlier_iep = lapply(baselink_earlier_full,
                         function(x) fread(x))

years_middle = seq(from = 2007, to = 2011)
baselink_middle_full = sprintf("%s%s.csv",
                                baselink_earlier_root,
                                years_middle)
all_middle_iep = lapply(baselink_middle_full,
                         function(x) fread(x))

### clean headers of data
head(all_earlier_iep[[1]])

clean_earlier_iep = lapply(all_earlier_iep,
                           function(x) { 
  x <- x %>%
    row_to_names(row_number = 5)
  
  return(x)
})


clean_middle_iep = lapply(all_middle_iep,
                           function(x) { 
  x <- x %>%
    row_to_names(row_number = 4)
  
  return(x)
})

### clean data and select pertinent columns of 2005-2011
filtered_earlier_iep = lapply(clean_earlier_iep, 
                               function(x) {
                                 x <- x %>%
                                   filter(Disability == "All disabilities") %>%
                                   dplyr::select(Year, State, "Age 3 to 21" = "Age 3-21 Totals")
                                 return(x)
                               })

filtered_middle_iep = lapply(clean_middle_iep, 
                               function(x) {
                                 x <- x %>%
                                   filter(Disability == "All disabilities")
                                 return(x)
                               })

names(filtered_middle_iep[[1]]) <- gsub("[\r\n]",
                            "", names(filtered_middle_iep[[1]]))
filtered_first_middle_iep <- filtered_middle_iep[[1]] %>%
  rename("Age 3 to 21" = "Ages3_21Total ") %>%
  dplyr::select(Year, State, "Age 3 to 21")

filtered_rest_middle_iep <- lapply(filtered_middle_iep[2:5], 
                               function(x) {
                                 x <- x %>%
                                   dplyr::select(Year, State, "Age 3 to 21")
                                 return(x)
                               })

filtered_earlier_iep_bound <- do.call(rbind.data.frame,
                                filtered_earlier_iep)

filtered_rest_middle_iep_bound <- do.call(rbind.data.frame,
                                filtered_rest_middle_iep)

filtered_earlier_middle_iep_bound = rbind(filtered_earlier_iep_bound, filtered_first_middle_iep, filtered_rest_middle_iep_bound)

## repeat for later links
baselink_later_root = "https://www2.ed.gov/programs/osepidea/618-data/state-level-data-files/part-b-data/child-count-and-educational-environments/bchildcountandedenvironments"
years_later = seq(from = 2012, to = 2017)

baselink_later_full_1 = sprintf("%s%s.csv",
                                baselink_later_root,
                                years_later[years_later < 2017])
baselink_later_full_2 = sprintf("%s%s-%s.csv",
                                baselink_later_root,
                                years_later[years_later >= 2017],
                               gsub("20", 
                                    "", 
                                    years_later[years_later >= 2017] +1))

all_later_iep = lapply(c(baselink_later_full_1,
                         baselink_later_full_2),
                         function(x) fread(x))

### clean headers of data
clean_later_iep = lapply(all_later_iep,
                           function(x) { 
  x <- x %>%
    row_to_names(row_number = 5) %>%
    clean_names()
  return(x)
})

clean_2016_iep <- all_later_iep[[5]] %>%
    row_to_names(row_number = 4) %>%
    clean_names() %>%
    dplyr::select(year, state_name, sea_education_environment, sea_disability_category, age_3_to_5, ages_6_21)

clean_later_iep[[5]] <- clean_2016_iep
filtered_later_iep = lapply(clean_later_iep,
                           function(x) { 
  x <- x %>%
    dplyr::select(year, state_name, sea_education_environment, sea_disability_category, age_3_to_5, ages_6_21) %>%
    filter(sea_disability_category == "All Disabilities") %>%
    filter(grepl("Total",sea_education_environment))%>%
    mutate(age_3_to_5=replace(age_3_to_5, age_3_to_5 == "-", 0)) %>%
    mutate(ages_6_21=replace(ages_6_21, ages_6_21 == "-", 0)) %>%
    group_by(year, state_name) %>%
    summarise("Age 3 to 21" = sum(as.numeric(age_3_to_5), as.numeric(ages_6_21))) %>%
    rename(State = state_name, Year = year)
  return(x)
})

filtered_later_iep_bound <- do.call(rbind.data.frame,
                                filtered_later_iep)

iep_init = rbind(data.frame(filtered_earlier_middle_iep_bound), data.frame(filtered_later_iep_bound))

nonstates = paste(toupper(unique(c(grep("Republic", unique(iep_init$State), value = TRUE, ignore.case = TRUE), 
              grep("Outlying", unique(iep_init$State), value = TRUE, ignore.case = TRUE),
              grep("Puerto|Federated|Marianas|BIE|Samoa|50 states|Bureau|Guam|Virgin Island", unique(iep_init$State), 
                   value = TRUE, ignore.case = TRUE)))), collapse = "|") 

iep_analytic <- iep_init %>%
  rename_all(toupper) %>%
  mutate(STATE = toupper(STATE), # since half states are lowercase; others are uppercase
    AGE_3_21_TOTAL = as.numeric(gsub(",", "", AGE.3.TO.21))) %>%
  dplyr::select(YEAR, STATE, AGE_3_21_TOTAL) %>%
  filter(!grepl(nonstates, STATE)) #filtering out those outside 50 states = DC


## check to see if # of unique states equal to 51 
## 50 states + DC
stopifnot(length(unique(iep_analytic$STATE)) == 51)
```


## Write to intermed analytic to file

```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

fwrite(iep_analytic,
"../intermediate_objects/iepcounts_allstates_20052018.csv")

```

# Step three: construct two rates and summarize 

- Count of due process hearing *requests* by state divided by total student enrollment. You'll want to look at the documentation for the complaints data to get the right var/you can ping me on slack when the data is clean, but I think the varname is something like DPCTotal (which should be a higher count than the variable DPC adjudicated which references filings that go to a hearing)
- Count of due process hearing *requests* by state divided by total number of students aged 3-21 with IEP

- Show DC first compared to median of other states
- Then compare to a few other high complaint east coast states (e.g., NY)

```{r}
## before merges, check overlap
diffstates = setdiff(unique(dispute_analytic$STATE),
        unique(totalenroll_analytic$STATE))
diffyears = setdiff(unique(dispute_analytic$FALL_SY),
        unique(totalenroll_analytic$YEAR))
if(length(diffstates) != 0){
    print(sprintf("Data dispute analytic has following states not in enrollment",
                  paste(diffstates, collapse = ";")))
        }



dph_tse <- merge(dispute_analytic, totalenroll_analytic, by.x=c('FALL_SY','STATE'),
                 by.y = c("YEAR", "STATE"),
                 all.x = TRUE) 


diffstates = setdiff(unique(dph_tse$STATE),
        unique(iep_analytic$STATE))
diffyears = setdiff(unique(dph_tse$YEAR),
        unique(iep_analytic$YEAR))
dph_iep <- merge(dph_tse, iep_analytic, by=c('YEAR','STATE')) %>%
        mutate(dph_iep_rate = ((DUE_PROCESS_COMPLAINTS_DPC_TOTAL + EXPEDITED_DUE_PROCESS_COMPLAINTS_EDPC_TOTAL)/AGE_3_21_TOTAL)*10000,
              dph_tse_rate = 
            ((DUE_PROCESS_COMPLAINTS_DPC_TOTAL+EXPEDITED_DUE_PROCESS_COMPLAINTS_EDPC_TOTAL)/STUDENT_COUNT)*10000,
              mediation_iep_rate = (MEDIATION_REQUESTS_TOTAL/AGE_3_21_TOTAL)*10000,
              wsc_iep_rate = (WRITTEN_SIGNED_COMPLAINTS_WSC_TOTAL/AGE_3_21_TOTAL)*10000,
               STATE = case_when(
              STATE == "DISTRICT OF COLUMBIA" ~ "DC",
              TRUE ~ stringr::str_to_title(STATE)
               ))



## state groups
dc <- "DC"
east = c(dc, "New York", "Virginia", "Maryland",
         "New Jersey",
         "Texas")




scaleFUN <- function(x) sprintf("%.1f", x)

dc_col = "#E91436"
ggplot(dph_iep %>%
      filter(STATE %in% east), 
  aes(x = YEAR, y = dph_iep_rate,
      group = STATE)) +
  geom_point(color = dc_col, size = 3) +
  geom_line(color = dc_col, size = 1) +
  facet_wrap(~STATE, scales = "free") +
  expand_limits(y = 0) +
  ylab("Count of due process filings\nper 10,000 IEP students") +
  theme_new() +
  theme(axis.text.x = element_text(angle = 90,
                                   size = 14)) +
  xlab("School year (fall)") +
  scale_y_continuous(breaks = pretty_breaks(n = 8))

ggsave("../output/dphrate_periep.pdf",
       plot = last_plot(),
       device = "pdf",
       width = 12,
       height = 8)

ggplot(dph_iep %>%
      filter(STATE %in% east), 
  aes(x = YEAR, y = dph_tse_rate,
      group = STATE)) +
  geom_point(color = dc_col, size = 3) +
  geom_line(color = dc_col, size = 1) +
  facet_wrap(~STATE, scales = "free") +
  expand_limits(y = 0) +
  ylab("Count of due process filings\nper 10,000 students") +
  theme_new() +
  theme(axis.text.x = element_text(angle = 90,
                                   size = 14)) +
  xlab("School year (fall)") +
  scale_y_continuous(breaks = pretty_breaks(n = 8),
                     labels = scaleFUN)


ggsave("../output/dphrate_perstud.pdf",
       plot = last_plot(),
       device = "pdf",
       width = 12,
       height = 8)


## then, show DC relative to the median
## medians
dph_nondc = dph_iep %>%
        filter(!STATE %in% dc) %>%
        group_by(YEAR) %>%
        summarise_at(c("dph_iep_rate", 
                             "mediation_iep_rate",
                             "wsc_iep_rate"),
                     median, na.rm = TRUE) %>%
        mutate(STATE = "Median\n(excluding DC)") 
dph_dc = dph_iep %>%
      filter(STATE == dc) %>%
      dplyr::select(STATE, YEAR, 
                    dph_iep_rate,
                    mediation_iep_rate,
                    wsc_iep_rate)
dph_nymaryland = dph_iep %>%
      filter(STATE %in% grep("New York|Maryland", east, value = TRUE)) %>%
      dplyr::select(STATE, YEAR, 
                    dph_iep_rate,
                    mediation_iep_rate,
                    wsc_iep_rate)

dph_comparetypes = rbind.data.frame(dph_nondc,
                                    dph_dc,
                                    dph_nymaryland) %>%
                reshape2::melt(, id.vars = c("STATE", "YEAR")) %>%
                mutate(disp_type = 
                      case_when(grepl("^dph", variable) ~ "DPH\nfilings",
                                grepl("^med", variable) ~ "Mediation\nrequests",
                                grepl("wsc", variable) ~ "Written\nsigned complaints"))



ggplot(dph_comparetypes, aes(x = YEAR, y = value,
                             fill = STATE,
                             group = STATE)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~disp_type, scales = "free_y", ncol = 1) +
  theme_new() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, size = 14)) +
  ylab("Per 10,000 IEP students") +
  xlab("School year (fall)") +
  labs(fill = "") + 
  guides(fill = guide_legend(ncol = 2)) +
  scale_fill_manual(values = 
                  c("DC" = dc_col,
                    "Median\n(excluding DC)" = "wheat4",
                    "New York" = "#002D72",
                    "Maryland" = "#EAAB00"))


ggsave("../output/type_variation.pdf",
       plot = last_plot(),
       device = "pdf",
       width = 8,
       height = 12)

```
