---
title: "Trends in due process hearings over time"
author: "Rebecca Johnson (raj2@princeton.edu)"
date: "5/5/2019"
output: html_document
---

```{r}
library(data.table)
library(ggplot2)
library(dplyr)
library(devtools)
library("jsonlite")
library(stringr)
library(scales)
library(xtable)
library(stringr)

#devtools::install_github('UrbanInstitute/education-data-package-r')


theme_new <- function(base_size = 24, base_family = "Helvetica"){
  theme_bw(base_size = base_size, base_family = base_family) %+replace%
    theme(
      panel.grid = element_blank(),   
      panel.border = element_rect(fill = NA, colour = "black", size=1),
      panel.background = element_rect(fill = "white", colour = "black"), 
      strip.background = element_rect(fill = NA),
      axis.text.x = element_text(color = "black"),
      axis.text.y = element_text(color = "black")
      )
}

```


# Step one: load and merge numerator data on complaints by state

General link: https://www2.ed.gov/programs/osepidea/618-data/state-level-data-files/index.html#part-b-menu

## Pull all the years from that site

```{r}
start_fall = 2006
start_spring= 2007
end_fall = 2017
end_spring = 2018

falls = seq(from = start_fall, to = end_fall, by = 1)
springs = seq(from = start_spring, to =  end_spring, by = 1)
spring_suffix = gsub("20", "", springs)
urls_dr = sprintf("https://www2.ed.gov/programs/osepidea/618-data/state-level-data-files/part-b-data/dispute-resolution/bdispres%s-%s.csv",
                  falls,
                  spring_suffix)

## iterate and read in
all_dr_years = list()
for(i in 1:length(urls_dr)){
  one_dr_year = fread(urls_dr[i])
  all_dr_years[[i]] = one_dr_year
}

## remove empty rows
all_dr_years_removeblank = lapply(all_dr_years, function(x){x[rowSums(x == "") == 0]})

## subset to actual data columns
all_dr_years = lapply(all_dr_years_removeblank,
                                function(x){x[2:nrow(x), ]})
all_dr_years_colnames = lapply(all_dr_years_removeblank, 
                               function(x){unlist(x[1, ])})
all_dr_years_colnames_n = lapply(all_dr_years_colnames, 
                                 length)

## go df by df and try to merge

### first see which don't match

for(i in 1:length(all_dr_years_colnames)){
  print(sprintf("difference between df %s and df %s cols: ",
                i, i+1))
  print(setdiff(unlist(all_dr_years_colnames[i]),
                unlist(all_dr_years_colnames[i+1])))
  
}


### for RG: stopped here; reconcile colnames
### until they match and you can rbind
### make sure all up to 2017-2018 are captured
### ping me if you have questions about how to reconcile the names 
### (think it's mainly removing weird suffixes like  1.2)

all_dr_years_firstwo_renamed = lapply(all_dr_years[1:2],
                              setNames,
                            unique(unlist(all_dr_years_colnames[1:2])))

all_dr_years_firstwo_bound = do.call(rbind.data.frame,
                                     all_dr_years_firstwo_renamed)

all_dr_years_7through11_renamed = lapply(all_dr_years[7:11],
                              setNames,
                          unique(unlist(all_dr_years_colnames[7:11])))
all_dr_years_7through11_bound = do.call(rbind.data.frame,
                                all_dr_years_7through11_renamed)
table(all_dr_years_7through11_bound$Year)
table(all_dr_years_firstwo_bound$Year)

## rg: this code initially just used later years
## bc it was too much work to reconcile columns
## ideally, would create an analytic dataset
## with all years pulled above
## after colnames have been reconciled
df_analytic_prerename = all_dr_years_7through11_bound
colnames(df_analytic_prerename) =   gsub("\\s+|\\(|\\)", 
                          "_", 
                        colnames(all_dr_years_7through11_bound)) 
df_analytic_rename_2 = df_analytic_prerename
colnames(df_analytic_rename_2) = gsub("[0-9]+.*$|__",
                            "",
                            colnames(df_analytic_prerename))

df_analytic = df_analytic_rename_2
```

## Write complaints data to file

For RG: updated path to write to so it's stored in same github repo

```{r}

### set working directory so that it's where
### current file is stored
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

fwrite(df_analytic,
"../intermediate_objects/complaints_allstates.csv")



```

# Step two: load and merge data on denominator (total count of students and count of students with ieps)

## Denom 1: total count of students

Original data source: NCES, Common core data, elsi table generator: https://nces.ed.gov/ccd/elsi/tableGenerator.aspx


```{r}
totalenroll_raw = read.csv("../raw_input/elsi_enrollment_removeheaderfooter.csv")

## clean:
## 1. use total students all grades excludes AE as the correct count
## 2. might want to reshape from its current wide format
## to long format (states repeated across school years)
## to make easier to auto-calculate rates

## then, write clean copy to tje intermediate objects directory

```

## Denom 2: count of students with IEP (IDEA part B)

Get from same site as complaints data since not in NCES data

```{r}
## link for sy 2005-2006 to 2011-2012
baselink_earlier_root = "https://www2.ed.gov/programs/osepidea/618-data/state-level-data-files/part-b-data/child-count/bchildcount"
years_earlier = seq(from = 2005, to = 2011)
baselink_earlier_full = sprintf("%s%s.csv",
                                baselink_earlier_root,
                                years_earlier)
all_earlier_iep = lapply(baselink_earlier_full,
                         function(x) fread(x))

## for RG: see that need to do cleaning  of headers to get  the actual columns (seem to start at row 5)
head(all_earlier_iep[[1]])

## repeat for later links
baselink_later_root = "https://www2.ed.gov/programs/osepidea/618-data/state-level-data-files/part-b-data/child-count-and-educational-environments/bchildcountandedenvironments"
years_later = seq(from = 2012, to = 2018)

### see from looking at links that it's 
### just the fall of the year up through 2016
### then it switches to whole year
baselink_later_full_1 = sprintf("%s%s.csv",
                                baselink_later_root,
                                years_later[years_later < 2017])
baselink_later_full_2 = sprintf("%s%s-%s.csv",
                                baselink_later_root,
                                years_later[years_later >= 2017],
                               gsub("20", 
                                    "", 
                                    years_later[years_later >= 2017] +1))

all_later_iep = lapply(c(baselink_later_full_1,
                         baselink_later_full_2),
                         function(x) fread(x))


## clean each and then rowbind
## main column to retain is the name of the state,
## the year,
## and the count of iep students, which is
## Age 3-21 totals in earlier data
## and separated into 3-5 and 6-21
## in later data, so you might need to find sum
```


# Step three: construct two rates and summarize 

- Count of due process hearing *requests* by state divided by total student enrollment. You'll want to look at the documentation for the complaints data to get the right var/you can ping me on slack when the data is clean, but I think the varname is something like DPCTotal (which should be a higher count than the variable DPC adjudicated which references filings that go to a hearing)
- Count of due process hearing *requests* by state divided by total number of students aged 3-21 with IEP

- Show DC first compared to median of other states
- Then compare to a few other high complaint east coast states (e.g., NY)


## Older code that might be relevant but refers to older data / varnames

Might be useful but doesn't run as is due to changes in the input data!


```{r}


## merge using state name
colnames(compare_states) = gsub("\\s+", "_", colnames(compare_states))
compare_states = compare_states %>%
            rename(total_students = `Total_Students_[State]_2013-14`,
                   total_districts = `Total_Number_Operational_School_Districts_[District]_2013-14`)
unique(compare_states$State_Name)
unique(dph$State)


dph_mainyear = dph %>%
      filter(Year == "2016-2017") %>%
      mutate(dph_filed = as.numeric(gsub(",", "", as.character(`Due_Process_ComplaintsDPCTotal`))),
             dph_hearings = as.numeric(gsub(",", "", 
                              as.character(`DPC_Hearingsfully_adjudicatedTotal`)))) %>%
      dplyr::select(State, dph_filed, dph_hearings) %>%
      mutate(State_Name = toupper(State)) %>%
      dplyr::select(-State) 
      

## merge

compare_w_dph = merge(compare_states,
                      dph_mainyear,
                      by = "State_Name",
                      all.x = TRUE) %>%
            left_join(ieps %>% dplyr::select(State_Name, ncesiepcount),
                      on = "State_Name") %>%
            mutate(ncesiepcount = ifelse(State_Name == "DISTRICT OF COLUMBIA",
                                         10644, ncesiepcount)) %>%
            mutate(filings_perstudent = (dph_filed/total_students)*10000,
                   hearings_perstudent = (dph_hearings/total_students)*10000,
                   ieps_perstudent= (ncesiepcount/total_students)*100,
                   students_per_district = total_students/total_districts,
                   state_category = case_when(State_Name %in% c("DISTRICT OF COLUMBIA",
                                                                "TEXAS",
                                                                "OHIO") ~ "Case study",
                                    State_Name %in% states_obtained ~ "Filings obtained",
                                    TRUE ~ "Filings not obtained"),
                   state_title = str_to_title(State_Name),
                   hearing_per_filing = ifelse(dph_filed == 0,
                                          0, dph_hearings/dph_filed))

ggplot(compare_w_dph, aes(x = reorder(state_title, filings_perstudent), 
                          y = filings_perstudent,
                          color = state_category)) +
  geom_point(size = 4) +
  ylab("Due process filings per\n10,000 students") +
  theme_new(base_size = 24) +
  theme(axis.text.y = element_text(size = 10),
        legend.position = c(0.7, 0.2),
        legend.background = element_blank()) +
  scale_color_manual(values = c("#FDE725FF", 
                                "wheat4",
                              "#238A8DFF")) +
  coord_flip() +
  labs(color = "") +
  xlab("")

ggsave("../output/dph_rate.pdf",
       plot = last_plot(),
       width = 9,
       height = 12,
       device = "pdf")


## create dc specific plot
focus_states = compare_w_dph %>% filter(State_Name %in% c("DISTRICT OF COLUMBIA", "VIRGINIA", "MARYLAND", "NEW YORK"))
national_median = quantile(compare_w_dph$filings_perstudent,
                           probs = 0.5)
focus_states_forplot = rbind.data.frame(focus_states %>%
                                      dplyr::select(state_title,
                                                    filings_perstudent,
                                                    state_category),
                          data.frame(state_title = "National\nmedian",
                                     filings_perstudent = national_median,
                                     state_category = "Filings obtained")) %>%
                          mutate(clean_state = 
                                      case_when(state_title == "District Of Columbia" ~
                                                "DC",
                                                TRUE ~ state_title),
                                 clean_category = 
                                  case_when(clean_state == "DC" ~ "Case study",
                                            TRUE ~ "Other"))
ggplot(focus_states_forplot, aes(x = reorder(clean_state, filings_perstudent), 
                          y = filings_perstudent,
                          fill = clean_category)) +
  geom_bar(stat = "identity") +
  ylab("Due process filings per\n10,000 students") +
  theme_new(base_size = 24) +
  theme(legend.position = c(0.7, 0.2),
        legend.background = element_blank()) +
  scale_fill_manual(values = c("Case study" = "#132460",
                                "Other" = "#999999")) +
  coord_flip() +
  labs(color = "") +
  xlab("") +
  geom_label(aes(x = reorder(clean_state, filings_perstudent),
                 y = filings_perstudent,
                  label = round(filings_perstudent, 2)),
             fill = "white", size = 14) +
  guides(fill = FALSE)  +
  ylim(-5, 50)

ggsave("../output/dph_rate_dcfocus.jpg",
       plot = last_plot(),
       width = 9,
       height = 12,
       device = "jpg")
  

ggplot(compare_w_dph %>%
      filter(state_title != "Hawaii"),
      aes(x = reorder(state_title, students_per_district), 
                          y = students_per_district,
                          color = state_category)) +
  geom_point(size = 4) +
  ylab("Total students divided by total\nnumber of districts") +
  theme_new(base_size = 24) +
  scale_y_continuous(breaks = pretty_breaks(n = 10)) +
  theme(axis.text.y = element_text(size = 10),
        legend.position = c(0.7, 0.2),
        axis.text.x = element_text(angle = 90),
        legend.background = element_blank()) +
  scale_color_manual(values = c("#FDE725FF", 
                                "wheat4",
                              "#238A8DFF")) +
  coord_flip() +
  labs(color = "") +
  xlab("")

ggsave("../output/dist_consolidation.pdf",
       plot = last_plot(),
       width = 9,
       height = 12,
       device = "pdf")

ggplot(compare_w_dph, aes(x = reorder(state_title, ieps_perstudent), 
                          y = ieps_perstudent,
                          color = state_category)) +
  geom_point(size = 4) +
  ylab("Percent of students who have\nIEPs") +
  theme_new(base_size = 24) +
  scale_y_continuous(breaks = pretty_breaks(n = 10)) +
  theme(axis.text.y = element_text(size = 10),
        legend.position = c(0.3, 0.8),
        legend.background = element_blank()) +
  scale_color_manual(values = c("#FDE725FF", 
                                "wheat4",
                              "#238A8DFF")) +
  coord_flip() +
  labs(color = "") +
  xlab("")

ggsave("../output/iep_rate.pdf",
       plot = last_plot(),
       width = 9,
       height = 12,
       device = "pdf")







## 
dph_states_wdata = dph %>%
          filter(State %in% filings_progress$state)



dist = read.csv("../data/allstates/district_consolidation.csv") %>%
      mutate(n_students = )

dph_1314 = dph %>%
  filter(Year == "2013-2014") %>%
  left_join(statedisp %>% filter(springschoolyear == 2014) %>% %>% mutate(state = as.character(state)) %>%
                    dplyr::select(state, ncesiepcount),
                    by = "state")




filings_progress = filings_progress %>%
          left_join(dph %>% 
                      filter(Year == "2012-2013") %>%
                      dplyr::select(State, Due_Process_ComplaintsDPCTotal),
                    by = c("state" = "State")) %>%
          left_join(statedisp %>% filter(springschoolyear == 2013) %>% mutate(state = as.character(state)) %>%
                    dplyr::select(state, ncesiepcount),
                    by = "state") %>%
        left_join(dist, by = "state") %>%
          mutate(n_districts = as.numeric(gsub(",", "", n_districts)), 
                 students_per_district = as.numeric(gsub(",", "", avg_per_district)),
                 complaints_total = as.numeric(Due_Process_ComplaintsDPCTotal),
                 ieps_total = ifelse(state != "District of Columbia",
                                  as.numeric(ncesiepcount), 
                                  11000),
              complaints_per_100iep = (complaints_total/ieps_total)*100) %>%
            dplyr::select(-avg_per_district, -excluded_districts, -n_schools,
                          -Due_Process_ComplaintsDPCTotal, -ncesiepcount, -n_students)
cols_select = c("state", "years", "format", "n_districts",
                "students_per_district",
                "complaints_total", "ieps_total", "complaints_per_100iep")
print(xtable(filings_progress[, cols_select]), include.rownames = FALSE)

```

